Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-02 sections/eval.tex
Update on Overleaf.

\subsubsection{Finch Logic}
The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor. This means that operators take in a set of indexed tensors and output an indexed tensor. Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order. To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$. A $\finchquery$ assigns the output of an expression to a name. We use this to denote that we materialize an expression. Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation. A plan is a sequence of queries and outputs a set of tensors.

@@

We now describe how to define a few example functions from the API in this language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$. For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
\footnotesize &\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
\footnotesize &\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
\footnotesize &\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}
Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops. In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels. However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.

>>>>

\subsubsection{Finch Logic}
The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor. This means that operators take in a set of indexed tensors and output an indexed tensor. Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order. To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$. A $\finchquery$ assigns the output of an expression to a name. We use this to denote that we materialize an expression. Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation. A plan is a sequence of queries and outputs a set of tensors.

@@

We now describe how to define a few example functions from the API in this language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$. For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}
Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops. In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels. However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.
